---
title: "Applied Logistic Regression Homework 4"
output: html_document
---

##Exercise One

For this exercise, you will use the Hyponatremia dataset (this is the same data from week 4). You can download the hyponatremia.dta Stata file, or you can also access the data through this [CSV file](https://d396qusza40orc.cloudfront.net/logisticregression/data/hyponatremia.csv) .
```{r}
hypNa <- read.csv("C:/Users/Karmen/Desktop/All Things Coursera/current/Applied Logistic Regression/datasets/hyponatremia.csv")
str(hypNa)
```

If you recall from last week, the data for this homework exercise derive from an epidemiological study of hyponatremia (a life-threatening condition) among runners of the 2002 Boston Marathon. Hyponatremia is defined as an electrolyte disturbance in which the serum sodium concentration is lower than normal (<135 mmol/l).


Complete the following:

**a)Assess the association between hyponatremia (dichotomous variable nas135) and sex (variable female) by making a 2 by 2 table. Calculate the odds ratio of hyponatremia of a female compared to a male. Compute the 95% confidence interval for this odds ratio. Interpret the findings.   **   
`nas135` is a dichotomous variable where 1 indicates present hyponatremia, and 0 indicates absent hyponatremia.
```{r}
table(hypNa$nas135, hypNa$female)
(ORfm <- (37 * 297)/ (25 * 129)) 
```
To calculate odds ratio from a table:
$$\widehat{OR} = \frac{odds\ hyponatremia\ in\ female}{hyponatremia\ in\ male} = \frac{\frac{hyponatremic\ female}{not\ hyponatremic\ female}}{\frac{hyponatremic\ male}{not\ hyponatremic\ male}}$$
$$\widehat{OR} =\frac{hyponatremic\ female}{not\ hyponatremic\ female} * \frac{not\ hyponatremic\ male}{hyponatremic\ male}$$
$$\widehat{OR} = \frac{37}{129}*\frac{297}{25}$$
$$\widehat{OR} = 3.407442$$
Based on this calculation, odds for women to lack sodium are 3.4 times higher than odds for males. 
Since $\widehat{OR}$ is not normally distributed and 
$$\widehat{OR} = e^{\beta_1}$$
and 
$$\ln\left(\widehat{OR}\right) = \beta_1$$
where natural logarithm of $|idehat{OR}$ is more normally distributed, we'll use it to calculate 95% confidence interval for $\widehat{OR}$
So, 
$$\beta_1 = \ln(3.407442) = 1.225962$$
We can find variance of $\beta_1$ and then its standard error from formula:
$$\widehat{Var}(\beta_1 ) = \left[\frac{1}{a} + \frac{1}{b} +\frac{1}{c} + \frac{1}{d}\right]$$
and standard error of $\beta_1$ is
$$\widehat{SE}(\beta_1 ) = \left[\frac{1}{a} + \frac{1}{b} +\frac{1}{c} + \frac{1}{d}\right]^{\frac{1}{2}}$$
$$\widehat{SE}(\beta_1 ) = \left[\frac{1}{297} + \frac{1}{129} +\frac{1}{25} + \frac{1}{37}\right]^{\frac{1}{2}}$$
$$\widehat{SE}(\beta_1) = 0.279546$$
Now we can calculate 95% confidence interval for $\beta_1$ as:
$$95\% CI(\beta_1) = \beta_1 \pm 1.96 * \widehat{SE}(\beta_1)$$
$$95\% CI(\beta_1) = 1.225962 \pm 1.96 * 0.279546$$
$$95\% CI(\beta_1) = (0.6780518, 1.7738722)$$
Then 95% CI for $\widehat{OR}$ is
$$95\% CI(\widehat{OR}) = (e^{0.6780518}, e^{1.7738722})$$
$$95\% CI(\widehat{OR}) = (1.970036, 5.893631)$$

**b)Perform a logistic regression analysis with Stata using nas135 as dependent variable and female as the only independent variable. Use the Likelihood Ratio test to assess the significance of the model. Is the model with female a better model than the naïve model? **    

Performing logistic regression in R with `nas135` as the dependent variable and `female` as the only covariate:
```{r}
logfit <- glm(nas135 ~ female, data = hypNa, family = binomial(link = "logit"))
summary(logfit)
```
As we can see, $\beta_1$ and $\widehat{SE}(\beta_1)$ match with the ones calculated by hand. Now to assess whether our model with `female` as predictor for lack of sodium is better than the naive model with no predictors:
```{r}
#likelihood ratio:
(LR <- logfit$null.deviance - logfit$deviance)

#degrees of freedom for the likelihood ratio:
(df <- logfit$df.null - logfit$df.residual)

(pValue <- pchisq(LR, df = df, lower.tail = FALSE))
```

**If you are using Stata, you can use their built-in statistical functions to obtain p-values (type help functions).**     
**c)What is the naïve model? What is the probability of hyponatremia that this model predicts?**   
Naive model is model with no predictors. We can simply make a table for variable `nas135`   
```{r}
(tbl <- table(hypNa$nas135))
prop.table(tbl) * 100
```
Naive model predicts 12.70492 % chances for hyponatremia for each person, regardless of their gender - that is  the percent of hyponatremiacs in this dataset.   

**d)Run a logistic regression analyses with no independent variables. Transform the coefficient obtained from this model into a probability.**   


```{r}
naive_logfit <- glm(nas135 ~ 1, data = hypNa, family = binomial(link = "logit"))
summary(naive_logfit)
```
Since 
$$Probability = \frac{Odds}{1 + Odds}$$
and we can calculate odds so to exponentiate the only coefficient we get from the output
$$Probability = \frac{e^{-1.927305 }}{1 + e^{-1.927305 }}$$
and we get
$$Probability =0.1270492$$
```{r}
(beta_naive <- naive_logfit$coefficient)

unname(exp(beta_naive)/(1 + exp(beta_naive)))
```


**e)Using the model with female as independent variable, compute the estimated probability of hyponatremia per males and females. Write down the equation for the logit.**

First, I'll add new variable to the dataset that captures both genders and it's a factor variable, then I'll make logistic regresion with that variable as predictor - that should a dummy variable for gender by default.
```{r}
hypNa$gender <- as.factor(hypNa$female)
logfit2 <- glm(nas135 ~ gender, data = hypNa, family = binomial(link = "logit"))
summary(logfit2)
```
Equation for the logit :

$$g(gender) = \beta_0 + \beta_1 * gender$$  
for females(gender = 1) we have
$$g(gender = 1) = \beta_0 + \beta_1 * 1 = -1.248894 $$
and for males(gender = 0) we have
$$g(gender = 0) = \beta_0 + \beta_1 * 0 = \beta_0 = -2.4749$$

```{r}
(logit_fem <- coef(logfit2)[1] + coef(logfit2)[2]*1)
(logit_male <- -2.4749)

#probability for females:
(Prob_fem <- exp(logit_fem)/(1 + exp(logit_fem)))

#probability for males
(Prob_male <- exp(logit_male) / (1 + exp(logit_male)))


(pred_male <- predict(logfit2, newdata = data.frame(gender = as.factor(0)), type = "response"))
(pred_female <- predict(logfit2, newdata = data.frame(gender = as.factor(1)), type = "response"))
```
We predict that, based on our model with gender as predictor for hyponatremia, females have on average 22.29% chances to experience hyponatremia, while males have only 7.76% chances for same.

**f)Use the Wald test to assess the significance of the coefficient for female.**
```{r}
coefs <- summary(logfit2)$coefficient
coefs
(Wald <- coefs[2,1] / coefs[2,2])
(pValueW <- pnorm(Wald, lower.tail = FALSE))
```
p-value for the Wald test is highly significant, which means that gender is a significant predictor for hyponatremia.

**g)Fit a model with runtime as the only independent variable. Assess the significance of the model.**
```{r}
log_runtime <- glm(nas135 ~ runtime, data = hypNa, family = binomial(link = "logit"))
summary(log_runtime)
(coefs <- summary(log_runtime)$coefficients)

(Wald_r <- coefs[2,1] / coefs[2,2])
(pValue_r <- pnorm(Wald_r, lower.tail = FALSE))
```
p-value of Wald test for model with only `runtime` as predictor is highly significant.     

**h)Calculate the probability of hyponatremia of a runner who takes 4 hours (240 minutes) to complete the marathon.**
```{r}
(runner240 <- predict(log_runtime, newdata = data.frame(runtime = 240), type = "response"))
```
Runner who takes 4 hours to complete the marathon has 13.33 %  chances to develop hyponatremia.    

**i)Fit a model with female and runtime as independent variables. Assess the significance of the model. Which null hypothesis is tested?**
```{r}
log_fr <- glm(nas135 ~ female + runtime, data = hypNa, family = binomial(link = "logit"))
summary(log_fr)
coefs <- summary(log_fr)$coefficients
(LR <- log_fr$null.deviance <- log_fr$deviance)
(dfLR <- log_fr$df.null - log_fr$df.residual)

(pValueLR <- pchisq(LR, df = dfLR, lower.tail = FALSE))
```
In this hypothesis test the null hypothesis is that both $\beta_1$ and $\beta_2$ are both equal to zero, whereas the alternative hypothesis says that at least one of them is not zero. I've done likelihood ratio test that compares model with 2 predictors with the naive model without predictors. The p value from that test came out to be highly significant.  